---
title: "Marriage Licenses Over Time"
subtitle: "A plot of marriage licenses over time"
author: 
  - Aman Rana
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

data <- read_csv(here::here("data/analysis_data/marriage_analysis_data.csv"))
```


# Introduction

We use @citeR, @opendatatoronto, and @rohan.

This paper is to show marriage licenses over time in Toronto. @sec-data

# Data {#sec-data}

```{r}
#| label: fig-marriage
#| fig-cap: Marriage Licenses Issued
#| echo: false

data |> 
  ggplot(aes(x = date, y = marriage_licenses)) +
  geom_point()

```

We see marriages have seasonality in @fig-marriage as expected. 
Data becomes sparser post 2020, we can assume this could be from effects from covid-19, lower appetite for marriage in tougher socio-economic environments. But it could be a feature of the data collection process itself. It is worth reaching out to @opendatatoronto to see if its because more recent marriage license data is unavailable due to some underlying governmental process.

\newpage


# References


